---
title: "Indeed Data Science Research"
author: "Team DAREZ"
date: "10/5/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
require(rvest)
require(readr)
require(tidyverse)
require(DT)
require(odbc)
require(xml2)
```


```{r}
library(odbc)
my_connection <- dbConnect(drv = odbc::odbc(),
Driver = "SQL Server",
                 server = "data607.database.windows.net",
         database = "Project2",
         uid = "project2admin",
         pwd = "DATA607login")
```


## Scraping Indeed

```{r}
#Specifying the url for desired website to be scraped
# url <- 'https://www.indeed.com/jobs?q=data%20scientist&l&vjk=00ba1a22ba67ffd2'

#Reading the HTML code from the website
#webpage <- read_html(url)

#Using CSS selectors to scrape the rankings section
#job_data_html <- html_nodes(webpage,'.jobtitle , #sja0 b') 
# The characters '.text-primary' came from selector gadget

#Converting the ranking data to text
#job_data <- html_text(job_data_html)

#Let's have a look at the rankings
#head(job_data)
```


Sample Test

```{r}
url <- 'https://www.indeed.com/jobs?q=data%20scientist&l&vjk=00ba1a22ba67ffd2'
webpage <- read_html(url)
job_data_html <- html_nodes(webpage,'.jobtitle , #sja0 b') 
job_data <- html_text(job_data_html)
head(job_data)
```
```{r}
str_extract(job_data, "(\\w+.+)+") 
```

```{r}
url <- 'https://www.indeed.com/jobs?q=data%20scientist&l&vjk=00ba1a22ba67ffd2'
webpage <- read_html(url)
job_data_html <- html_nodes(webpage,'.salaryText') 
job_data <- html_text(job_data_html)
head(job_data)
```
```{r}
# First set of digits that are formatted like sal per year
# AKA Min Salary of the range
MinSal_data <- str_extract(job_data, "\\$\\d+,\\d+ ")
MinSal_data <- str_remove_all(MinSal_data, " |\\$")
# Max salary of the range
MaxSal_data <- str_extract(job_data, "- \\$\\d+,\\d+ ")
MaxSal_data <- str_remove_all(MaxSal_data, "- \\$| ")
# salary_data <- str_extract(job_data, "(\\d+.+)+") 
salary_data <- str_remove_all(job_data, "\\d+.+ an hour| a year|\\$")
salary_data <- str_remove_all(salary_data, "\\n")
# salary_data <- str_remove_all(salary_data, " a year")
# salary_data <- str_remove_all(salary_data,"\\$")
salary_data
MinSal_data
MaxSal_data
```


```{r}
salary_data <- str_remove_all(salary_data, "\\d+ an hour")
salary_data <- str_remove_all(salary_data, " a year")
salary_data <- as.data.frame(salary_data)
salary_data[2,1] <- ("70,000 - $90,000")
salary_data %>% 
  mutate(salary = as.factor(salary_data)) %>%
  count(salary)
```


```{r}
url <- 'https://www.indeed.com/jobs?q=data%20scientist&l&vjk=00ba1a22ba67ffd2'
webpage <- read_html(url)
date_data_html <- html_nodes(webpage,'.date') 
date_data <- html_text(date_data_html)
head(date_data)
```
```{r}
date_data
str_extract(date_data, "\\d+|Just posted|Today")
```

  #get the job title
  job_title <- page %>% 
    rvest::html_nodes("div") %>%
    rvest::html_nodes(xpath = '//a[@data-tn-element = "jobTitle"]') %>%
    rvest::html_attr("title")
    
```{r}
empties <- data.frame(title=character(),
                    date=character(),
                 company=character(),
                  salary=character(),
                    maxsal=character(),
                  minsal=character(),
                 location=character(), 
                  summary=character(), 
                    link=character(), 
                  # description = character(),
                 stringsAsFactors=FALSE) 
for (i in seq(0, 800, 10)){
  url <- paste0('https://www.indeed.com/jobs?q=data%20scientist&start=10&vjk=c03d341a13e2bb74',i)
  var <- read_html(url)
  # job title    
    title <-  var %>% 
        html_nodes('#resultsCol .jobtitle') %>%
          html_text() %>%
        str_extract("(\\w+.+)+") 
# job posting date
    date <-  var %>% 
        html_nodes('.date') %>%
          html_text() %>%
        str_extract("\\d+|Just posted|Today") 
# employer name
    company <- var %>% 
        html_nodes('#resultsCol .company') %>%
          html_text() %>%
        str_extract("(\\w+).+") 
# salary ranges
    job_data <- var %>%
        html_nodes('.salaryText') %>%
          html_text()
            salary <- str_remove_all(job_data, "\\d+.+ an hour| a year|\\n|\\$")
            maxsal <- str_extract(job_data, "- \\$\\d+,\\d+ ") %>%
              str_remove_all("- \\$| ")
            minsal <- str_extract(job_data, "\\$\\d+,\\d+ ") %>%
              str_remove_all(" |\\$")
# job location
    location <- var %>%
        html_nodes('#resultsCol .location') %>%
          html_text() %>%
        str_extract("(\\w+.)+,.[A-Z]{2}") 
# job summary
    summary <- var %>%
        html_nodes('#resultsCol .summary') %>%
          html_text() %>%
        str_extract(".+")
# link to the job description
    link <- var %>%
        html_nodes('#resultsCol .jobtitle .turnstileLink, #resultsCol a.jobtitle') %>%
          html_attr('href') 
            link <- paste0("https://www.indeed.com",link)

  Indeed <- rbind(empties, as.data.frame(cbind(title,
                                                  date,
                                                  company,
                                                  salary,
                                                  maxsal,
                                                  minsal,
                                                  location,
                                                  summary,
                                                  link)))
}
```



```{r}
Indeed
```



```{r}
write.csv(listings, "C:/bigdata/Indeed.csv")
```

    
    

```{r}
empties <- data.frame(title=character(),
                    date=character(),
                  company=character(), 
                 salary=character(),
                 maxsal=character(),
                 minsal=character(),
                 location=character(), 
                 summary=character(), 
                 links=character(),
                 stringsAsFactors=FALSE) 
for (i in seq(0, 50, 10)){
  url <- paste0('https://www.indeed.com/jobs?q=data%20scientist&start=10&vjk=c03d341a13e2bb74',i)
  var <- read_html(url)
  
  
title <-  var %>% 
      html_nodes('.jobtitle, #sja0 b') %>%
    html_text() %>%
      str_extract("(\\w+.+)+") 
# date <-  var %>% 
#      html_nodes('.date') %>%
#    html_text() %>%
#      str_extract("\\d+|Just posted|Today") 
company <- var %>% 
      html_nodes('.company') %>%
    html_text() %>%
      str_extract("(\\w+).+") 
#job_data <- var %>%
#    html_nodes('.salaryText') %>%
#    html_text()
#      salary <- str_remove_all(job_data, "\\d+.+ an hour| a year|\\n|\\$")
#      maxsal <- str_extract(job_data, "- \\$\\d+,\\d+ ") %>%
#       str_remove_all("- \\$| ")
#      minsal <- str_extract(job_data, "\\$\\d+,\\d+ ") %>%
#    str_remove_all(" |\\$")
location <- var %>%
        html_nodes('.location') %>%
      html_text() %>%
        str_extract("(\\w+.)+,.[A-Z]{2}")   
summary <- var %>%
        html_nodes('.summary') %>%
      html_text() %>%
        str_extract(".+")
links <- var %>%
        html_nodes('.jobtitle .turnstileLink, a.jobtitle') %>%
      html_attr('href') 
        links <- paste0("https://www.indeed.com",links)
        
Indeed <- rbind(empties, as.data.frame(cbind(title,
                                                  #date,
                                                  company,
                                                  #salary,
                                                  #maxsal,
                                                  #minsal,
                                                  location,
                                                  summary,
                                                  links
                                                  )))
}
```





```{r}
Indeed
```




```{r}
write.csv(listings, "C:\bigdata")
```




  #get the company name
  company_name <- page %>% 
    rvest::html_nodes("span")  %>% 
    rvest::html_nodes(xpath = '//*[@class="company"]')  %>% 
    rvest::html_text() %>%
    stringi::stri_trim_both() -> company.name 
  
  
  #get job location
  job_location <- page %>% 
    rvest::html_nodes("span") %>% 
    rvest::html_nodes(xpath = '//*[@class="location"]')%>% 
    rvest::html_text() %>%
    stringi::stri_trim_both()


```{r}
page_result_start <- 10 # starting page 
page_result_end <- 30 # last page results
page_results <- seq(from = page_result_start, to = page_result_end, by = 10)

full_df <- data.frame()
for(i in seq_along(page_results)) {
  first_page_url <- "https://www.indeed.com/jobs?q=data%20scientist&l&vjk=68011ffc792c1cc8"
  url <- paste0(first_page_url, "&start=", page_results[i])
  page <- xml2::read_html(url)
  Sys.sleep(2)
  
  # get links
  links <- page %>% 
    rvest::html_nodes("div") %>%
    rvest::html_nodes(xpath = '//*[@data-tn-element="jobTitle"]') %>%
    rvest::html_attr("href")
  
  job_description <- c()
  for(i in seq_along(links)) {
    
    url <- paste0("https://ca.indeed.com/", links[i])
    page <- xml2::read_html(url)
    
    job_description[[i]] <- page %>%
      rvest::html_nodes("span")  %>% 
      rvest::html_nodes(xpath = '//*[@class="jobsearch-JobComponent-description icl-u-xs-mt--md"]') %>% 
      rvest::html_text() %>%
      stringi::stri_trim_both()
  }
  df <- data.frame(job_description)
  full_df <- rbind(full_df, df)
}
```


## Cleaning that data

```{r}
# strsplit(title_data, "\n")
```




```{r}
data <- read_delim(job_data, "\t", quote = "\\\"", 
    escape_double = FALSE, col_names = FALSE, 
    na = "empty", trim_ws = TRUE)
```



# From Example

Create an empty data frame called listings with the type of character()


```{r}
empties <- data.frame(title=character(),
                    date=character(),
                  company=character(), 
                 salary=character(),
                 maxsal=character(),
                 minsal=character(),
                 location=character(), 
                 summary=character(), 
                 links=character(),
                 stringsAsFactors=FALSE) 
for (i in seq(0, 20, 10)){
  url_ds <- paste0('https://www.indeed.com/jobs?q=data%20scientist&l&vjk=dd25f8809ed80778',i)
  var <- read_html(url_ds)
  Sys.sleep(3)
title <-  var %>% 
      html_nodes('.jobtitle, #sja0 b') %>%
    html_text() %>%
      str_extract("(\\w+.+)+") 
date <-  var %>% 
      html_nodes('.date') %>%
    html_text() %>%
      str_extract("\\d+|Just posted|Today") 
company <- var %>% 
      html_nodes('.company') %>%
    html_text() %>%
      str_extract("(\\w+).+") 
job_data <- var %>%
    html_nodes('.salaryText') %>%
    html_text()
      salary <- str_remove_all(job_data, "\\d+.+ an hour| a year|\\$")
      maxsal <- str_extract(job_data, "- \\$\\d+,\\d+ ") %>%
       str_remove_all("- \\$| ")
      minsal <- str_extract(job_data, "\\$\\d+,\\d+ ") %>%
    str_remove_all(" |\\$")
location <- var %>%
        html_nodes('.location') %>%
      html_text() %>%
        str_extract("(\\w+.)+,.[A-Z]{2}")   
summary <- var %>%
        html_nodes('.summary') %>%
      html_text() %>%
        str_extract(".+")
links <- var %>%
        html_nodes('.jobtitle .turnstileLink, a.jobtitle') %>%
      html_attr('href') 
        link <- paste0("https://www.indeed.com",link)
        
Indeed <- rbind(empties, as.data.frame(cbind(title,
                                                  date,
                                                  company,
                                                  salary,
                                                  maxsal,
                                                  minsal,
                                                  location,
                                                  summary,
                                                  links
                                                  )))
}
```

```{r}
links_descs <- data.frame(links=character(),
                    descriptions=character())

for (i in seq(0, 20, 10)){
  url_ds <- paste0('https://www.indeed.com/jobs?q=data%20scientist&l&vjk=dd25f8809ed80778',i)
  var <- read_html(url_ds)
  Sys.sleep(3)
  #links
  links <- var %>%
    html_nodes('.jobtitle .turnstileLink, a.jobtitle') %>%
    html_attr('href') 
    link <- paste0("https://www.indeed.com",link)
  
  for(i in seq_along(link)) {
    url <- paste0("https://www.indeed.com/", link[i])
    full_desc <- read_html(url) 

    description[[i]] <- full_desc %>%
      html_nodes('.jobsearch-ViewJobLayout-innerContentGrid')  %>% 
      html_text() %>% 
    }
  descriptions <- rbind(links_descs, as.data.frame(cbind(links,
                                                   descriptions
                                                  )))
}
```




```{r}
  links <- str_remove_all(links,"^ w+")
  url <- paste0("https://www.indeed.com", links[2])
  
  
```



```{r}
datatable(descriptions)
```


Bringing it all together in a table

```{r}
datatable(listings)
```



```{r}
head(listings$links, 3)
```


