---
title: "HW10"
author: "Zachary Palmore"
date: "4/7/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

Smith is in jail and has 1 dollar; he can get out on bail if he has 8 dollars. A guard agrees to make a series of bets with him. If Smith bets A dollars, he wins A dollars with probability .4 and loses A dollars with probability .6.

## Exercises

Find the probability that he wins 8 dollars before losing all of his money if

(a) he bets 1 dollar each time (timid strategy).



```{r}
p <- 0.4
q <- 0.6
N <- 8 
timidP <- signif(((q/p)^1 - 1 ) / ((q/p)^8 - 1 ), 3)
timidP
```



(b) he bets, each time, as much as possible but not more than necessary to bring his fortune up to 8 dollars (bold strategy).


```{r}
boldP <- p^3
boldP
```



(c) Which strategy gives Smith the better chance of getting out of jail? 

The bold strategy give Smith the best chance of getting out of jail with a probability of `r boldP` compared to `r timidP` with the timid strategy. However, both chances are terrible given the circumstances, should not be recommended. 
