---
title: "Module1"
author: "Zachary Palmore"
date: "9/2/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F)
```

**Principles of Data Visualization and Introduction to ggplot2**

I have provided you with data about the 5,000 fastest growing companies in the US, as compiled by Inc. magazine. lets read this in:

```{r}
inc <- read.csv("https://raw.githubusercontent.com/charleyferrari/CUNY_DATA_608/master/module1/Data/inc5000_data.csv", header= TRUE)
```

And lets preview this data:

```{r}
head(inc)
```



```{r}
summary(inc)
```


Think a bit on what these summaries mean. Use the space below to add some more relevant non-visual exploratory information you think helps you understand this data:

```{r,}
# Insert your code here, create more chunks as necessary
library(tidyverse)
library(ggpubr)
library(psych)
theme_set(theme_minimal()) # Set plot theme
sum(is.na(inc)) # 12 missing values
inc[which(is.na(inc)),] # Print all missing value indecies
# They hold no significance and can be removed completely
inc.omitted <- na.omit(inc) # remove missing values; equivalent to complete.cases() but worked up front
inc.ccs <- inc[complete.cases(inc),] # To show equivalency
```



```{r, results='hide'}
# Only 8 variables; we can plot all
inc %>% 
  dplyr::select(-Name, -City, -State, -Industry) %>% # 4 characters variables are useless in plot
  gather(key, value) %>% # gather into key value pairs
  ggplot(aes(value)) + # create ggplot
  geom_boxplot() + # as a geometric boxplot
  facet_wrap(~key, scales ="free") + # by each key
  theme(axis.ticks.x = element_blank()) # hide x axis tick marks

# Results:
# There are a lot of outliers
# Comparing other variables to rank which has no outliers
# The remaining variables are dominated by outliers
```



```{r}
# Look at NY
inc %>% 
  filter(State == "NY") %>% 
  group_by(Industry) %>% 
  summarise(IndMed = median(Employees)) # Calc median employment by industry in this state

# Look at how industries are represented
inc %>% 
  dplyr::count(Industry, sort=T)

# Look closer at statistics
describe(inc)
# Growth Rate, Revenue, and the number of Employees have high skew values
# Those same values are also curtailed sharply mid distribution
# These are to be expected in the fastest growing 5000 companies
```



```{r results='hide', fig.show='hide'}
# Out of curiosity
# Which states did best in their ranking? 
inc %>% 
  arrange(desc(Rank)) %>% 
  group_by(State) %>% 
  summarise(StateRank = (sum(Rank)/nrow(inc))) %>%
  ggplot(aes(reorder(State, StateRank), StateRank)) + 
  geom_col(aes(fill = StateRank, alpha = .80)) + coord_flip() +
  labs(y = "Averaged Cumulative State Rank", x = "State", 
       title = "Highest Ranked States from Fastest 5000 Companies", caption = "Data compiled and ranked by Inc. magazine") +
  theme(legend.position = "none", 
        panel.grid.minor.x = element_line(color = "lightgrey",
                                          linetype = "dotted"),
        panel.grid.minor.y = element_line(color = "lightgrey", 
                                          linetype = "dotted"),
        plot.title = element_text(hjust = 0.5),
        plot.caption = element_text(hjust = 0.5))
# Could have also used a histogram with stat = "count" 
```



## Question 1

Create a graph that shows the distribution of companies in the dataset by State (ie how many are in each state). There are a lot of States, so consider which axis you should use. This visualization is ultimately going to be consumed on a 'portrait' oriented screen (ie taller than wide), which should further guide your layout choices.


```{r, fig.width=6, fig.height=8}
# Answer Question 1 here
data.frame(table(inc$State)) %>% 
  ggplot(aes(y = reorder(Var1, Freq), x = Freq)) + 
  geom_col(aes(fill = Freq, alpha = .80, width = .9)) + 
  labs(x = "State", y = "Count of \"Fastest\" Companies", 
    title = "Number of Fastest Growing Companies",
    caption = "Data on fastest 5000 companies compiled by Inc.") +
  geom_text(aes(label = round(Freq, 0)), size = 3.6, hjust = 1) +
  theme(legend.position = "none", 
        panel.grid.minor.x = element_line(color = "lightgrey",
                                          linetype = "dotted"),
        panel.grid.minor.y = element_line(color = "lightgrey", 
                                          linetype = "dotted"),
        plot.title = element_text(hjust = 0.5),
        plot.caption = element_text(hjust = 0.5), 
        axis.ticks.x = element_blank(), 
        axis.ticks.y = element_blank(),
        axis.text.y = element_text(size = 10), 
        axis.text.x = element_blank(), 
        axis.title.x = element_text(size = 10), 
        axis.title.y = element_blank(),) 
```





## Quesiton 2

Lets dig in on the state with the 3rd most companies in the data set. Imagine you work for the state and are interested in how many people are employed by companies in different industries. Create a plot that shows the average and/or median employment by industry for companies in this state (only use cases with full data, use R's `complete.cases()` function.) In addition to this, your graph should show how variable the ranges are, and you should deal with outliers.

```{r, results='hide'}
# This sounds like a boxplot/violin plot given an average/median employment by industry with a variable range for each state but would also show outliers unless removed 
# Method of removing outliers? None specified. 
# We view several cases to see the full picture to decide 
library(ggpubr)
# Let's first view it pre-outlier removal

# Find state to filter by
data.frame(table(inc$State)) %>% 
  arrange(desc(Freq))
# Based on this table of frequencies; NY is 3rd 

# Create boxplot with complete cases and all outliers shown
box.inc.ccs <- inc.ccs %>% 
  filter(State == "NY") %>%
  ggplot(aes(x = reorder(Industry, Employees, median), Employees)) + 
  geom_boxplot(aes(fill = median(Employees), alpha = .75)) + coord_flip() + 
  stat_summary(fun.y=median, geom="point", shape=21, 
               size=2, color="lightblue", alpha = 0.9, fill = "white") + 
  labs(y = "Employees", x = "Industry", 
       title = "NY Employment by Industry", 
       subtitle = "Median shown as light blue dot for each industry",
       caption = "From data on fastest 5000 companies as compiled by Inc. magazine") +
  theme(legend.position = "none", 
        panel.grid.minor.x = element_line(color = "lightgrey",
                                          linetype = "dotted"),
        panel.grid.minor.y = element_line(color = "lightgrey", 
                                          linetype = "dotted"),
        plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        plot.caption = element_text(hjust = 0.5),
        axis.ticks.x = element_blank(), 
        axis.text.y = element_text(face = "bold", size = 12), 
        axis.text.x = element_text(size = 12), 
        axis.title.x = element_text(face = "bold", size = 12), 
        axis.title.y = element_blank())  + 
  scale_x_discrete(limits = rev(levels(inc$Var1)), expand = c(-0.8, -4))  +
  annotate("text", x = 3, y = 2000, label = "No Outliers Removed")

# Now create boxplot with two clearest outliers not shown
box.inc.tworemoved <- inc.ccs %>% 
  filter(State == "NY") %>%
  ggplot(aes(x = reorder(Industry, Employees, median), Employees)) + 
  geom_boxplot(aes(fill = median(Employees), alpha = .75)) + coord_flip() + 
  stat_summary(fun.y=median, geom="point", shape=21, 
               size=2, color="lightblue", alpha = 0.9, fill = "white") + 
  labs(y = "Employees", x = "Industry", 
       title = "NY Employment by Industry", 
       subtitle = "Median shown as light blue dot for each industry",
       caption = "From data on fastest 5000 companies as compiled by Inc. magazine") +
  theme(legend.position = "none", 
        panel.grid.minor.x = element_line(color = "lightgrey",
                                          linetype = "dotted"),
        panel.grid.minor.y = element_line(color = "lightgrey", 
                                          linetype = "dotted"),
        plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        plot.caption = element_text(hjust = 0.5),
        axis.ticks.x = element_blank(), 
        axis.text.y = element_text(face = "bold", size = 12), 
        axis.text.x = element_text(size = 12), 
        axis.title.x = element_text(face = "bold", size = 12), 
        axis.title.y = element_blank())  + 
  scale_x_discrete(limits = rev(levels(inc$Var1)), expand = c(-0.8, -4))  +
  annotate("text", x = 3, y = 600, label = "Two Outliers Removed") + 
  scale_y_continuous(limits = c(0, 1000), breaks = seq(0, 1000, 500), expand = c(0.01, 0.5)) 

# Alternative options for removal and clarity
# Remove outliers based on IQR of R boxplot
outs <- boxplot(inc.ccs$Employees, plot=F)$out
outs.num <- outs %>% as.numeric() %>% data.frame()
outs.num %>%
  count() # 610 Outliers Identified via Boxplot
# Down select to remove them
inc.boxremoval <- inc.ccs[-which(inc.ccs$Employees %in% outs),]

# Applying the above method on average employment per industry we have this
# Boxplot with outliers determined by 1.5 times each industy's IQR removed
box.inc.boxremoval <- inc.boxremoval %>% 
  filter(State == "NY") %>%
  ggplot(aes(x = reorder(Industry, Employees, median), Employees)) + 
  geom_boxplot(aes(fill = median(Employees), alpha = .75)) + coord_flip() + 
  stat_summary(fun.y=median, geom="point", shape=21, 
               size=2, color="lightblue", alpha = 0.9, fill = "white") + 
  labs(y = "Employees", x = "Industry", 
       title = "NY Employment by Industry", 
       subtitle = "Median shown as light blue dot for each industry",
       caption = "From data on fastest 5000 companies as compiled by Inc. magazine") +
  theme(legend.position = "none", 
        panel.grid.minor.x = element_line(color = "lightgrey",
                                          linetype = "dotted"),
        panel.grid.minor.y = element_line(color = "lightgrey", 
                                          linetype = "dotted"),
        plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        plot.caption = element_text(hjust = 0.5),
        axis.ticks.x = element_blank(), 
        axis.text.y = element_text(face = "bold", size = 12), 
        axis.text.x = element_text(size = 12), 
        axis.title.x = element_text(face = "bold", size = 12), 
        axis.title.y = element_blank())  + 
  scale_x_discrete(limits = rev(levels(inc$Var1)), expand = c(-0.8, -4)) + 
  annotate("text", x = 2, y = 164, label = "Outliers Removed beyond 1.5xIQR")
#  scale_y_continuous(limits = c(0, 200), breaks = seq(0, 200, 50), expand = c(0.01, 0.5)) + 

inc.boxremoval %>% 
  filter(State == "NY") %>% 
  filter(Industry == "Government Services") # Cipher Tech Solutions - only one in NY for "Government Services" industry
inc.boxremoval %>% 
  filter(State == "NY") %>% 
  filter(Industry == "Environmental Services") # Creative Environment Solutions and one other environmental service are the only two for "Environmental Services" industry
```


```{r, results='hide', fig.show='hide'}
ggarrange(box.inc.ccs, box.inc.tworemoved, box.inc.boxremoval)

inc.ccs %>% 
  filter(State == "NY") %>% 
  filter(Industry == "Consumer Products & Services") %>% 
  arrange(desc(Employees))
inc.ccs %>%
  filter(State == "NY") %>%
  filter(Industry == "Business Products & Services") %>%
  arrange(desc(Employees))
```



```{r, fig.dim=c(8,6)}
# Answer Question 2 here
inc.ccs %>% 
  filter(State == "NY") %>%
  ggplot(aes(x = reorder(Industry, Employees, median), Employees)) + 
  geom_boxplot(aes(fill = median(Employees), alpha = .75)) + coord_flip() + 
  stat_summary(fun.y=median, geom="point", shape=21, 
               size=1.6, color="lightblue", alpha = 0.9, fill = "white") + 
  labs(y = "Employees", x = "Industry", 
       title = "NY Employment by Industry", 
       subtitle = "Median shown as light blue dot for each industry",
       caption = "From data on fastest 5000 companies as compiled by Inc. magazine") +
  theme(legend.position = "none", 
        panel.grid.minor.x = element_line(color = "lightgrey",
                                          linetype = "dotted"),
        panel.grid.minor.y = element_line(color = "lightgrey", 
                                          linetype = "dotted"),
        plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        plot.caption = element_text(hjust = 0.5),
        axis.ticks.x = element_blank(), 
        axis.text.y = element_text(size = 11), 
        axis.text.x = element_text(size = 11), 
        axis.title.x = element_text(size = 11), 
        axis.title.y = element_blank())  + 
  scale_x_discrete(limits = rev(levels(inc$Var1)), expand = c(-0.8, -4))  +
  annotate("text", x = 5, y = 700, label = "Two Outliers Not Shown:") + 
  annotate("text", x = 3, y = 700, label = "Coty & Sutherland Global Services") +
  scale_y_continuous(limits = c(0, 1050), breaks = seq(0, 1000, 500), expand = c(0.01, 0.5)) 
```



## Question 3

Now imagine you work for an investor and want to see which industries generate the most revenue per employee. Create a chart that makes this information clear. Once again, the distribution per industry should be shown.

```{r, results='hide', fig.show='hide'}
# Testing Question 3 here
# There is more than one way to calculate "which industries generate the most revenue per employee"
# We take a look at NY out of curiosity for comparison
inc.ccs %>% 
  filter(State == "NY") %>%
  dplyr::select(Industry, Revenue, Employees) %>% 
  mutate(RevPerEmp = Revenue / Employees) %>%
  group_by(Industry) %>%
  summarise(TotRevPerEmp = (sum(RevPerEmp)/1000000)) %>% 
  ggplot(aes(reorder(Industry, TotRevPerEmp), TotRevPerEmp)) + 
  geom_col(aes(y = TotRevPerEmp, 
           fill = TotRevPerEmp, alpha = .80)) + coord_flip() + 
  labs(y = "Dollars (Millions)", x = "Industry", 
                      title = "Revenue Per Employee (NY)", 
                      caption = "Contains data on fastest 5000 companies as compiled by Inc. magazine") +
  theme(legend.position = "none", 
        panel.grid.minor.x = element_line(color = "lightgrey",
                                          linetype = "dotted"),
        panel.grid.minor.y = element_line(color = "lightgrey", 
                                          linetype = "dotted"),
        plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        plot.caption = element_text(hjust = 0.5))


# Then without NY filter (for entire US)
# Example 1: calculates the revenue per employee as a per capita dollar value in each company first 
# then sums the dollars values within each industry
inc.ccs %>% 
  dplyr::select(Industry, Revenue, Employees) %>% 
  mutate(RevPerEmp = Revenue / Employees) %>%
  group_by(Industry) %>%
  summarise(TotRevPerEmp = (sum(RevPerEmp)/1000000)) %>% 
  ggplot(aes(reorder(Industry, TotRevPerEmp), TotRevPerEmp)) + 
  geom_col(aes(y = TotRevPerEmp, 
           fill = TotRevPerEmp, alpha = .80)) + coord_flip() + 
  labs(y = "Dollars (Millions)", x = "Industry", 
                      title = "U.S. Revenue Per Employee", 
                      caption = "Contains data on fastest 5000 companies as compiled by Inc. magazine") +
  theme(legend.position = "none", 
        panel.grid.minor.x = element_line(color = "lightgrey",
                                          linetype = "dotted"),
        panel.grid.minor.y = element_line(color = "lightgrey", 
                                          linetype = "dotted"),
        plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        plot.caption = element_text(hjust = 0.5), 
        axis.ticks.x = element_blank(), 
        axis.text.y = element_text(face = "bold", size = 12), 
        axis.text.x = element_text(size = 12), 
        axis.title.x = element_text(face = "bold", size = 12), 
        axis.title.y = element_blank())  

# Scaled features
inc.ccs %>% 
  dplyr::select(Industry, Revenue, Employees) %>% 
  mutate(RevPerEmp = Revenue / Employees) %>%
  group_by(Industry) %>%
  summarise(TotRevPerEmp = (sum(RevPerEmp)/1000000)) %>% 
  ggplot(aes(reorder(Industry, TotRevPerEmp), TotRevPerEmp)) + 
  geom_col(aes(y = TotRevPerEmp, 
           fill = TotRevPerEmp, alpha = .80)) + coord_flip() + 
  labs(y = "Dollars (Millions)", x = "Industry", 
                      title = "U.S. Revenue Per Employee", 
                      caption = "Contains data on fastest 5000 companies as compiled by Inc. magazine") +
  theme(legend.position = "none", 
        panel.grid.minor.x = element_line(color = "lightgrey",
                                          linetype = "dotted"),
        panel.grid.minor.y = element_line(color = "lightgrey", 
                                          linetype = "dotted"),
        plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        plot.caption = element_text(hjust = 0.5), 
        axis.ticks.x = element_blank(), 
        axis.text.y = element_text(face = "bold", size = 12), 
        axis.text.x = element_text(size = 12), 
        axis.title.x = element_text(face = "bold", size = 12), 
        axis.title.y = element_blank()) + 
  scale_x_discrete(limits = rev(levels(inc$Var1)), expand = c(-0.8, -4))  +
  scale_y_continuous(limits = c(0, 200), breaks = seq(0, 200, 50), expand = c(0.01, 0.5))

# Example 2: calculates the total revenue as a dollar amount for each industry and 
# calculates separately the total number of employees in each industry 
# then creates a ratio of revenue dollars to the number of employees in each industry to find 
# the industry revenue per employee
inc.ccs %>% 
  dplyr::select(Industry, Revenue, Employees) %>% 
  group_by(Industry) %>%
  summarise(TotalRev = sum(Revenue),
            TotalEmp = sum(Employees),
            TotRevPerEmp = (TotalRev / TotalEmp)/1000) %>%
  ggplot(aes(reorder(Industry, TotRevPerEmp), TotRevPerEmp)) + 
  geom_col(aes(y = TotRevPerEmp, 
           fill = TotRevPerEmp, alpha = .80)) + coord_flip() + 
  labs(y = "Dollars (Thousands)", x = "Industry", 
                      title = "U.S. Revenue Per Employee", 
                      caption = "Contains data on fastest 5000 companies as compiled by Inc. magazine") +
  theme(legend.position = "none", 
        panel.grid.minor.x = element_line(color = "lightgrey",
                                          linetype = "dotted"),
        panel.grid.minor.y = element_line(color = "lightgrey", 
                                          linetype = "dotted"),
        plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        plot.caption = element_text(hjust = 0.5), 
        axis.ticks.x = element_blank(), 
        axis.text.y = element_text(face = "bold", size = 12), 
        axis.text.x = element_text(size = 12), 
        axis.title.x = element_text(face = "bold", size = 12), 
        axis.title.y = element_blank())  

# Example 3: calculates the revenue per employee as a per capita dollar value in each company 
# then finds the mean of those values for each industry 
inc.ccs %>% 
  dplyr::select(Industry, Revenue, Employees) %>% 
  mutate(RevPerEmp = Revenue / Employees) %>%
  group_by(Industry) %>%
  summarise(TotRevPerEmp = mean(RevPerEmp)/1000) %>% 
  ggplot(aes(reorder(Industry, TotRevPerEmp), TotRevPerEmp)) + 
  geom_col(aes(y = TotRevPerEmp, 
           fill = TotRevPerEmp, alpha = .80)) + coord_flip() + 
  labs(y = "Dollars (Thousands)", x = "Industry", 
                      title = "Revenue Per Employee (U.S)", 
                      caption = "Contains data on fastest 5000 companies as compiled by Inc. magazine") +
  geom_text(aes(label = round(TotRevPerEmp, 0)), size = 3, hjust = 1.19) +
  theme(legend.position = "none", 
        panel.grid.minor.x = element_line(color = "lightgrey",
                                          linetype = "dotted"),
        panel.grid.minor.y = element_line(color = "lightgrey", 
                                          linetype = "dotted"),
        plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        plot.caption = element_text(hjust = 0.5), 
        axis.ticks.x = element_blank(), 
        axis.text.y = element_text(face = "bold", size = 12), 
        axis.text.x = element_blank(),
        axis.title.x = element_text(face = "bold", size = 12), 
        axis.title.y = element_blank()) 
```


```{r}
# Answer Question 3 here
inc.ccs %>% 
  dplyr::select(Industry, Revenue, Employees) %>% 
  group_by(Industry) %>%
  summarise(TotalRev = sum(Revenue),
            TotalEmp = sum(Employees),
            TotRevPerEmp = (TotalRev / TotalEmp)/1000) %>%
  ggplot(aes(reorder(Industry, TotRevPerEmp), TotRevPerEmp)) + 
  geom_col(aes(y = TotRevPerEmp, 
           fill = TotRevPerEmp, alpha = .80)) + coord_flip() + 
  labs(y = "Dollars (Thousands)", x = "Industry", 
                      subtitle = "Fastest Growing Industries by Revenue Per Employee (U.S)", 
                      caption = "Contains data on fastest 5000 companies as compiled by Inc. magazine") +
  geom_text(aes(label = round(TotRevPerEmp, 0)), size = 3, hjust = 1.08) +
  theme(legend.position = "none", 
        panel.grid.minor.x = element_line(color = "lightgrey",
                                          linetype = "dotted"),
        panel.grid.minor.y = element_line(color = "lightgrey", 
                                          linetype = "dotted"),
        plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        plot.caption = element_text(hjust = 0.5), 
        axis.ticks.x = element_blank(), 
        axis.text.y = element_text(size = 11), 
        axis.text.x = element_blank(), 
        axis.title.x = element_text(size = 11), 
        axis.title.y = element_blank())   + 
  annotate("text", x = 3, y = 800, label = "Uses Total Dollar Value Ratios:", size = 3.5) + 
  annotate("text", x = 2, y = 800, 
           label = "Revenue Per Capita = sum(Revenue)/sum(Employees)", size = 3.5)
```



