---
title: "Module1"
author: "Group 3"
date: "9/2/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Principles of Data Visualization and Introduction to ggplot2**

I have provided you with data about the 5,000 fastest growing companies in the US, as compiled by Inc. magazine. lets read this in:

```{r}
inc <- read.csv("https://raw.githubusercontent.com/charleyferrari/CUNY_DATA_608/master/module1/Data/inc5000_data.csv", header= TRUE)
```

And lets preview this data:

```{r}
head(inc)
```



```{r}
summary(inc)
```


Think a bit on what these summaries mean. Use the space below to add some more relevant non-visual exploratory information you think helps you understand this data:

```{r}
# Insert your code here, create more chunks as necessary
library(tidyverse)
library(psych)
theme_set(theme_minimal()) # Set plot theme
sum(is.na(inc)) # 12 missing values
inc[which(is.na(inc)),] # Print all missing value indecies
# They hold no significance and can be removed completely
inc <- na.omit(inc) # remove missing values

# Only 8 variables; we can plot all
inc %>% 
  dplyr::select(-Name, -City, -State, -Industry) %>% # 4 characters variables are useless in plot
  gather(key, value) %>% # gather into key value pairs
  ggplot(aes(value)) + # create ggplot
  geom_boxplot() + # as a geometric boxplot
  facet_wrap(~key, scales ="free") + # by each key
  theme(axis.ticks.x = element_blank()) # hide x axis tick marks

# Results:
# There are a lot of outliers
# Comparing other variables to rank which has no outliers
# The remaining variables are dominated by outliers

# Look closer at statistics
describe(inc)
# Growth Rate, Revenue, and the number of Employees have high skew values
# Those same values are also curtailed sharply mid distribution
# These are to be expected in the fastest growing 5000 companies
```

```{r}
# Out of curiosity
# Which states did best in their ranking? 
inc %>% 
  arrange(desc(Rank)) %>% 
  group_by(State) %>% 
  summarise(StateRank = (sum(Rank)/nrow(inc))) %>%
  ggplot(aes(reorder(State, StateRank), StateRank)) + 
  geom_col(aes(fill = StateRank, alpha = .80)) + coord_flip() +
  labs(y = "Averaged Cumulative State Rank", x = "State", 
       title = "Highest Ranked States from Fastest 5000 Companies", caption = "Data compiled and ranked by Inc. magazine") +
  theme(legend.position = "none", 
        panel.grid.minor.x = element_line(color = "lightgrey",
                                          linetype = "dotted"),
        panel.grid.minor.y = element_line(color = "lightgrey", 
                                          linetype = "dotted"),
        plot.title = element_text(hjust = 0.5),
        plot.caption = element_text(hjust = 0.5))
# Could have also used a histogram with stat = "count" 
```



## Question 1

Create a graph that shows the distribution of companies in the dataset by State (ie how many are in each state). There are a lot of States, so consider which axis you should use. This visualization is ultimately going to be consumed on a 'portrait' oriented screen (ie taller than wide), which should further guide your layout choices.


```{r}
# Answer Question 1 here
data.frame(table(inc$State)) %>% 
  ggplot(aes(reorder(Var1, -Freq), Freq)) + 
  geom_col(aes(fill = Freq, alpha = .80)) + 
  coord_flip() + labs(x = "State", y = "Count of \"Fastest\" Companies", 
    subtitle = "Number of Fastest Growing Companies in each State",
    caption = "Data on fastest 5000 companies compiled by Inc.") +
  theme(legend.position = "none", 
        panel.grid.minor.x = element_line(color = "lightgrey",
                                          linetype = "dotted"),
        panel.grid.minor.y = element_line(color = "lightgrey", 
                                          linetype = "dotted"),
        plot.subtitle = element_text(hjust = 0.5),
        plot.caption = element_text(hjust = 0.5))
```

## Quesiton 2

Lets dig in on the state with the 3rd most companies in the data set. Imagine you work for the state and are interested in how many people are employed by companies in different industries. Create a plot that shows the average and/or median employment by industry for companies in this state (only use cases with full data, use R's `complete.cases()` function.) In addition to this, your graph should show how variable the ranges are, and you should deal with outliers.

```{r}
# This sounds like a boxplot/violin plot given an average/median employment by industry with a variable range for each state
# this would also show outliers unless removed
# However, the data is not averagable for the industry in the state
# Unless you did something like this 
inc %>% 
  filter(State == "NY") %>% 
  group_by(Industry) %>% 
  summarise(IndMed = median(Employees)) # Calc median employment by industry in this state

# Answer Question 2 here
data.frame(table(inc$State)) %>% 
  arrange(desc(Freq))
# Based on this table of frequencies; NY is 3rd 

# Remove outliers based on IQR of R boxplot
outs <- boxplot(inc$Employees, plot=F)$out
inc <- inc[-which(inc$Employees %in% outs),]

# Applying the above method on average employment per industry we have
inc %>% 
  filter(State == "NY") %>% 
  group_by(Industry) %>% 
  summarise(IndMed = median(Employees)) %>% 
  ggplot(aes(reorder(Industry, -IndMed)), IndMed) + 
  geom_col(aes(y = IndMed, fill = IndMed, alpha = .80)) + 
  coord_flip() + labs(y = "Average Number of Employees", x = "Industry", 
                      title = "Average Employment by Industry in NY", 
                      caption = "Contains data on fastest 5000 companies as compiled by Inc. magazine") +
  theme(legend.position = "none", 
        panel.grid.minor.x = element_line(color = "lightgrey",
                                          linetype = "dotted"),
        panel.grid.minor.y = element_line(color = "lightgrey", 
                                          linetype = "dotted"),
        plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        plot.caption = element_text(hjust = 0.5))


# Total employees per industry in the state of NY
inc %>% 
  filter(State == "NY") %>%
  group_by(Industry) %>%
  summarise(TotalEmployees = sum(Employees)) %>%
  ggplot(aes(reorder(Industry, -TotalEmployees)), TotalEmployees) + 
  geom_col(aes(y = TotalEmployees, fill = TotalEmployees, alpha = .80)) + 
  coord_flip() + labs(y = "Total Employees", x = "Industry", 
                      title = "NY Industry Employment", 
                      subtitle = "By Fastest Growing 5000 Companies") +
  theme(legend.position = "none", 
        panel.grid.minor.x = element_line(color = "lightgrey",
                                          linetype = "dotted"),
        panel.grid.minor.y = element_line(color = "lightgrey", 
                                          linetype = "dotted"),
        plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        plot.caption = element_text(hjust = 0.5))
```

## Question 3

Now imagine you work for an investor and want to see which industries generate the most revenue per employee. Create a chart that makes this information clear. Once again, the distribution per industry should be shown.

```{r}
# Answer Question 3 here
inc %>% 
  filter(State == "NY") %>%
  dplyr::select(Industry, Revenue, Employees) %>% 
  mutate(RevPerEmp = Revenue / Employees) %>%
  group_by(Industry) %>%
  summarise(TotRevPerEmp = (sum(RevPerEmp)/1000000)) %>% 
  ggplot(aes(reorder(Industry, -TotRevPerEmp), TotRevPerEmp)) + 
  geom_col(aes(y = TotRevPerEmp, 
           fill = TotRevPerEmp, alpha = .80)) + coord_flip() + 
  labs(y = "Revenue Per Employee (Millions of Dollars)", x = "Industry", 
                      title = "Total Revenue per Employee in NY", 
                      caption = "Contains data on fastest 5000 companies as compiled by Inc. magazine") +
  theme(legend.position = "none", 
        panel.grid.minor.x = element_line(color = "lightgrey",
                                          linetype = "dotted"),
        panel.grid.minor.y = element_line(color = "lightgrey", 
                                          linetype = "dotted"),
        plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        plot.caption = element_text(hjust = 0.5))
```




