---
title: "Exercise7_Post14"
author: "Zachary Palmore"
date: "4/20/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Prompt: 

Using R, provide the solution for any exercise in either Chapter 4 or Chapter 7 of the calculus textbook.  If you are unsure of your solution, post your concerns.

Exercise 7  Page 173  
Chapter 4.1

Section 4.1, Page 163, Exercise #7

Use 5 iterations of Newton’s Method with the given initial approximation to approximate the root. Compare it to the known value of the root.

```{r}
library(ggplot2)
library(knitr)
```

\[
f(x) = \ln(x), \ \ \ x_0 = 2
\]

Formula for Newton's Method: 

\[
x_{n+1} = x_n - \frac{f(x_n)}{f'(x_n)}
\]


```{r graph}
ggplot(data.frame(x = c(0, 10)), aes(x)) + 
  stat_function(fun = function(x) log(x)) +
  scale_x_continuous(breaks = c(1:10)) +
  geom_line(aes(y = 0))
```

\[ 
\dfrac{\mathrm{d}}{\mathrm{d}x}  = \frac{1}{x}
\]

```{r newton}
x0 <- 2
iters <- 5
newts <- c(0)
for(iter in 1:5){
  newt <- x0 - (log(x0)/(1/x0))
  newts[iter] <- newt
  x0 <- newt
}
kable(cbind(1:5, newts, 1, 1-newts), col.names = c("Iteration", "Newton Approximation", "Actual Value", "Difference"), align = "c")
```



```{r}
x <- 1.5
iter <- 5
newts <- c(0)
for(iter in 1:5){
  newt <- x - cos(x)/(1/(cos(x)))
  newts[iter] <- newt
  x <- newt
}
kable(cbind(1:5, newts, 1, 1-newts), col.names = c("Iteration", "Newton Approximation", "Actual Value", "Difference"), align = "c")
```



