---
title: "HW13"
author: "Zachary Palmore"
date: "4/29/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Question 1

Use integration by substitution to solve the integral below.

$$\int{4e^{-7x}dx}$$

## Answer 1 


$$-\frac{4}{7} e^{-7x} + C$$
Steps using substitution:

$U = -7x$, $dU = -7dx$, $dx = \frac{dU}{-7}$ so, 

$4\int e^U\frac{dU}{-7}$ = $\frac{4}{-7}\int e^UdU$ = $-\frac{4}{7} e^U + C$ thus, $-\frac{4}{7} e^{-7x} + C$

## Question 2

Biologists are treating a pond contaminated with bacteria. The level of contamination is changing at a rate of $\frac{dN}{dt} = -\frac{3150}{t^4}-220$ bacteria per cubic centimeter per day, where $t$ is the number of days since treatment began. Find a function $N(t)$ to estimate the level of contamination if the level after $1$ day was $6530$ bacteria per cubic centimeter.


## Answer 2

To find a function, take the integral of the rate:

$$\int {(\frac{-3150}{t^{4}}-220)dt}$$
$$\int {(-3150{t^{-4}}-220)dt}$$
$$\int-3150t^{-4}dt - \int -220dt$$
$$-3150\int t^{-4}dt - 220\int dt$$
$$-3150 \frac{-1}{3} t^{-3} - 220t$$
$$N(t)=  \frac{1050}{t^{3}}-220t+c$$



Solve for $c$ using the initial condition at day 1: $N(1)=6530$,

$$6530=  \frac{1050}{1^{3}}-220(1)+c$$ 
$$6530=1050-220+c$$
$$6530-1050+220=c$$
$$5700=c$$

Result: 

$$
N(t)=  \frac{1050}{t^{3}}-220t+5700
$$



## Question 3

Find the total area of the red rectangles in the figure below, where the equation of the line is $f(x)=2x-9$

## Answer 3 

Using the equation of the line and reading from the graph we estimate that the lower bound is 4.5 and upper is 8.5. Thus,

```{r}
library(stats) 
# If start and end points are truly 4.5 and 8.5 then
f <-function(x)(2*x-9)
integrate(f, lower = 4.5, upper = 8.5)
```

The area is 16. This is validated by counting the squares of the image which we assume has an area of 1 and width of 1.

## Question 4 

Find the area of the region bounded by the graphs of the given equations $y_1 = x_1^2 - 2x_1-2$ and $y_2 = x_2 + 2$.

Enter your answer below.

## Answer 4

Given:

$$y=x^{2}-2x-2$$
$$y=x+2$$

Solve for x to find the bounds:

$$x^{2}-2x-2=x+2$$
$$x^{2}-3x-4=0$$
$$(x-4)(x+1)=0$$
Thus, $x = -1$ and $x = 4$.

Using this equation we find the top and bottoms with integration. 

$$
\int _{ a }^{ b }{ (top-bottom) } dx
$$
So far we have $\int _{ -1 }^{ 4 }{ (x+2)-(x^{2}-2x-2) } dx$ which simplified becomes $\int _{- 1 }^{ 4 } {(-x^{2}+3x+4)dx}$. 

```{r}
## define the integrated function
integrand <- function(x)
  {
  -x^{2}+3*x+4
  }
## integrate the function from 0 to infinity
integrate(integrand, lower = -1, upper = 4)
```

Which we can use to calculate the area:

```{r}
area_fun <- function(x) {
  -x^2 + 3 * x + 4
}
integrate(area_fun, -1, 4)
```

The area is about 20.83.

## Question 5 

A beauty supply store expects to sell 110 flat irons during the next year. It costs $3.75 to store one flat iron for one year.There is a fixed cost of \$8.25 for each order. Find the lot size and the number of orders per year that will minimize 
inventory costs.


## Answer 5

Consider that $x$ is the number of irons to order and we have the following variables:

$$C_{order} -\ Cost\ per\ order$$
$$C_{year} -\ Yearly\ cost\ of\ ordering$$
$$R_{year} -\ Yearly\ cost\ of\ storage$$
$$R_{iron} - Storage\ rate\ of\ one\ flat\ iron\ for\ one\ year$$
$$n_{expected} -\ Expected\ number\ of\ orders\ in\ year$$
$$\bar{x}_{iron} - Average\ number\ of\ irons$$
$$I_{cost} =\ Cost\ of\ inventory$$

Our equations:
$$R_{year} = R_{iron} * \bar{x}_{iron}$$
$$C_{year} = C_{order}\frac{n_{expected}}{x}$$
$$I_{cost} = R_{year} + C_{order}$$

Which, when we enter what we have becomes:
$$3.75* \frac{x}{2} \hspace{8pt} or\ simply \hspace{8pt} 1.875*x$$
$$8.25 \times \frac{110}{x} \hspace{8pt} or\ simply \hspace{8pt} \frac{907.5}{x}$$
$$f(x) =1.875 * x + \frac{907.5}{x}$$
Calculate the minimized value by setting to zero and solving: 

$$f'(x) = 1.875-\frac{907.5}{x^2}$$
$$f'(x) = 0$$
$$1.875-\frac{907.5}{x^2} = 0$$
$$1.875 = \frac{907.5}{x^2}$$
$$1.875x^2 = 907.5$$
$$x^2 = \frac{907.5}{1.875}$$
$$x= \sqrt{\frac{907.5}{1.875}}$$
$$x=\sqrt{484}$$
$$x =22$$
$$quantity\ of\ irons\ = 22$$
$$110/22$$
$$quantity\ of\ orders\ = 5$$

The order should have 22 irons with a total of 5 orders. 


## Question 6 

Use integration by parts to solve the integral below.

$$\int{ln(9x) * x^6 dx}$$

## Answer 6 

Formula: $uv-\int vdu, u=ln(9x), du=\frac{1}{x}dx, dv=x^{6}, v=\frac{1}{7}x^{7}$

$$\frac{x^{7}ln(9x)}{7}-\int \frac{1}{7}x^{7}\frac{1}{x}dx$$
$$\frac{x^{7}ln(9x)}{7}-\frac{1}{7} \int \frac{x^{7}}{x}dx$$
$$\frac{x^{7}ln(9x)}{7}-\frac{1}{7}\int x^{6}dx$$
$$\frac{x^{7}ln(9x)}{7}-\frac{1}{7}(\frac{x^{7}}{7})+C$$
$$\frac{x^{7}ln(9x)}{7}-\frac{x^{7}}{49}+C$$

## Question 7 

Determine whether $f(x)$ is a probability density function on the interval $[1, e^6]$. If not, determine the value of the definite integral. 

$$f(x) = \frac{1}{6x}$$



## Answer 7 

$$\int_1^{e^6}\frac{1}{6x} dx = \frac{1}{6} ln(x)|_1^{e^6}$$
$$= \frac{1}{6} ln(e^6) - \frac{1}{6} ln(1)$$
$$= \frac{1}{6} \times 6 - \frac{1}{6} \times 0$$
$$= 1$$

The definite integral is 1 on the interval $[1, e^6]$.





