---
title: "HW15"
author: "Zachary Palmore"
date: "5/12/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## 1 

Find the equation of the regression line for the given points. Round any final values to the nearest hundredth, if necessary. 

( 5.6, 8.8 ), ( 6.3, 12.4 ), ( 7, 14.8 ), ( 7.7, 18.2 ), ( 8.4, 20.8 )

```{r}
x = c(5.6, 6.3, 7, 7.7, 8.4)
y = c(8.8, 12.4, 14.8, 18.2, 20.8)
reg <- lm(y~x)
reg 
# Equation of the line based on coefficients: -14.800 + 4.257x
plot(x,y, xlab="X", ylab="Y", main = "Regression Line of Points")
abline(reg)
lines(c(5,9), -14.800 + 4.257*c(5,9), col="aquamarine") 
text(6, 18, labels = "y = -14.800 + 4.257")
```

Rounded to the nearest hundreth the equation of the regression line is: 
$$y = -14.80 + 4.26x$$ 


## 2 

Find all local maxima, local minima, and saddle points for the function given below. Write your answer(s) in the form ( x, y, z ). Separate multiple points with a comma. 

$$f(x,y) = 24x-6xy^{2} - 8y^{3}$$

$$f_x = 24 - 6y^{2}$$
$$f_y = -12xy - 24y^{2}$$
$$- 6y^{2} + 24 = 0$$ 
$$-12xy -24y^{2} = 0$$

$$y = \sqrt{\frac{24}{6}} \pm2$$
$$x = -2y = 4$$

```{r}
fz <- function(x,y){
   z = 24*x-6*x*y^2-8*y^3
   print(paste("x =", x, ",", "y=", y, ",", "z=", z))
}
fz(-4,2)
```

The points separated with a comma in the format (x,y,z) are -4, 2, and -64. 

## 3

A grocery store sells two brands of a product, the "house" brand and a "name" brand. The manager estimates that if she sells the "house" brand for x dollars and the "name" brand for y dollars, she will be able to sell $81-21x+17y$ units of the "house" brand and $40+11x-23y$ units of the "name" brand. 

Step 1. Find the revenue function R ( x, y ).

$$R(x,y)=(81 - 21x + 17y)x + (40 + 11x - 23y)y$$
$$81x-21x^2+17xy+40y+11xy-23y^2$$
$$81x+40y+28xy-21x^2-23y^2$$


Step 2. What is the revenue if she sells the "house" brand for \$2.30 and the "name" brand for \$4.10?

$$R(2.3, 4.1)=81 * 2.3+40 * 4.1+28 * 2.3 * 4.1-21 * (2.3)^2-23 * (4.1)^2 = 116.62$$



## 4 

A company has a plant in Los Angeles and a plant in Denver. The firm is committed to produce a total of 96 units of a product each week. The total weekly cost is given by $C(x, y) = \frac{1}{6}x^2 + \frac{1}{6}y^2 + 7x + 25y + 700$, where x is the number of units produced in Los Angeles and y is the number of units produced in Denver. How many units should be produced in each plant to minimize the total weekly cost?

Given:
$C(x, y) = \frac{1}{6}x^2 + \frac{1}{6}y^2 + 7x + 25y + 700$

Where x is the units produced in LA and y units produced in Denver, we solve using the total units needed, 96, as shown:

$$\frac{1}{6}(96 - y)^2 + \frac{1}{6}y^2 + 7(96 - y) + 25y + 700$$
$$\frac{1}{6}(y^2 - 192y + 9216) + \frac{1}{6}y^2 + 18y + 25y + 1372$$
$$\frac{1}{3}y^{2} - 14y + 2908$$

Then find the minimum value:

$z = \frac{d}{dy}(\frac{1}{3}y^{2} - 14y + 2908$
$z = \frac{2}{3}y - 14$
$y = 21$

Substitute

$x = z - y$
$x = 96 - 21$
$x = 75$

We can also confirm this with the equations and local minima on plots of each location's unit production as shown below:

```{r}
par(mfrow=c(1,2))
curve(1/3*x^2-14*x+2908 , from = 0, to = 43, col="navy", main = "Denver Units")
curve(1/3*x^2-50*x+4636 , from = 0, to = 150, col="red", main = "LA Units")
```

It looks like our local minima are confirmed. Thus, for LA and Denver we have 75 units and 21 units respectively. 


## 5 

Evaluate the double integral on the given region.

$\int\int_R (e^{8x+3y}) dA, R:2\le x\le4\ and\ 2 \le y \le 4$

Write your answer in exact form without decimals. 

We have one form as: 

```{r}
1/24*((exp(32)+exp(16))*(exp(12) - exp(6)))
```

It could also be left in fractional form without decimals as mentioned:

$$\int_2^4\int_2^4 (e^{8x+3y})\ dy\ dx \int_2^4 (\frac{1}{3}e^{8x+3y})|_2^4\ dx$$
$$\int_2^4 ((\frac{1}{3}e^{8x+12})-(\frac{1}{3}e^{8x+6}))\ dx$$
$$\int_2^4 \frac{1}{3}e^{8x+6}(e^6-1)\ dx$$
$$\frac{1}{24}e^{38}(e^6-1)-\frac{1}{24}e^{22}(e^6-1)$$
$$\frac{1}{24}(e^6-1)(e^{38}-e^{22})$$
$$\frac{1}{24}(e^{22} - e^{28} - e^{38} + e^{44})$$

Without any decimals we have $\frac{1}{24}(e^{22} - e^{28} - e^{38} + e^{44})$