---
title: "HW7"
author: "Zachary Palmore"
date: "3/8/2021"
output:
  pdf_document: default
  html_document:
    df_print: paged
subtitle: "FUNDAMENTALS OF COMPUTATIONAL MATHEMATICS"
header-includes: 
- \newcommand{\bcenter}{\begin{center}}
- \newcommand{\ecenter}{\end{center}}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. 

Let X1, X2, . . . , Xn be n mutually independent random variables, each of
which is uniformly distributed on the integers from 1 to k. Let Y denote the
minimum of the Xi’s. Find the distribution of Y.


$$Y \hspace{4pt}over \hspace{4pt}the\hspace{4pt}interval\hspace{4pt} 1 \leq v \leq k, \hspace{4pt}f(v)\hspace{4pt} = \frac{(k - v + 1)^n - (k - v)^n}{k^n}$$

Where k is the number of possibilities in n mutually independent variables uniformly distributed from 1 to k, v the variables of at least one X, and f(v) the distribution function. 

```{r include=FALSE}
# https://math.dartmouth.edu/archive/m20f10/public_html/HW5Solutions.pdf
```


## 2. Your organization owns a copier (future lawyers, etc.) or MRI (future doctors). This machine has a manufacturer’s expected lifetime of 10 years. This means that we expect one failure every ten years. (Include the probability statements and R Code for each part.).


### a. What is the probability that the machine will fail after 8 years?. Provide also the expected value and standard deviation. Model as a geometric. (Hint: the probability is equivalent to not failing during the first 8 years..)

Modeled as an geometric distribution...

```{r}
y <- 8 # eighth year
p <- 1/10 # fails once every 10 years
pgeom(y, p, lower.tail = FALSE) # geometric probability of no failure in n years
pgeom(y, p, lower.tail = TRUE) # geometric probability of failure after n years
```


```{r}
# expected value
1/p
```


```{r}
# standard deviation 
round(sqrt((1-p)/(p^2)),4)
```

The probability that the machine will fail after 8 years is `r round(pgeom(y, p, lower.tail = FALSE), 4)`. 


### b. What is the probability that the machine will fail after 8 years?. Provide also the expected value and standard deviation. Model as an exponential. 

Modeled as an exponential distribution...

```{r}
y <- 8 # number of years
p = 1/10 # probability of failure
pexp(y, p, lower.tail = FALSE) # exponential probability of no failure in n years
pexp(y, p, lower.tail = TRUE) # exponential probability of failure after n years
```

```{r}
e <- 1/p # expected value
e
e # standard deviation
```

The probability that the machine will fail after 8 years is `r round(pexp(y, p, lower.tail = FALSE), 4)`. 



### c. What is the probability that the machine will fail after 8 years?. Provide also the expected value and standard deviation. Model as a binomial. (Hint: 0 success in 8 years) 

Modeled as a binomial distribution...

```{r}
y <- 8 # number of years
p = 1/10 # probability of failure
pbinom(0, y, p, lower.tail = FALSE)    # binomial probability of no failure in n years
pbinom(0, y, p, lower.tail = TRUE)  # binomial probability of failure after n years
```


```{r}
# expected value 
u <- y * p
u
```


```{r}
# standard deviation
sqrt(u * (1-p))
```

The probability that the machine will fail after 8 years is `r round(pbinom(0, y, p, lower.tail = TRUE), 4)`.



### d. What is the probability that the machine will fail after 8 years?. Provide also the expected value and standard deviation. Model as a Poisson. 


Modeled as a poisson distribution...

```{r}
y <- 8 # number of years
p = 1/10 # probability of failure
lam <- y * p
ppois(0, lam, lower.tail = FALSE)    # poisson probability of no failure in n years
ppois(0, lam, lower.tail = TRUE)   # poisson probability of failure after n years
```



```{r}
# expected value
lam
```


```{r}
# standard deviation
sqrt(lam)
```

The probability that the machine will fail after 8 years is `r round(ppois(0, lam, lower.tail = TRUE), 4)`.


```{r include=F}
# https://math.dartmouth.edu/archive/m20f10/public_html/HW5Solutions.pdf
# https://rpubs.com/raghu_74us/Assignment7
# https://rpubs.com/ofomicheva86/371898
# https://rpubs.com/Vishal0229/538609
# https://rstudio-pubs-static.s3.amazonaws.com/369146_0dcb8019640d4f65afce618d36d7d4e6.html
```












