---
title: "Lab 9"
author: "Zachary Palmore"
date: "`r Sys.Date()`"
output: openintro::lab_report
---

```{r load-packages, message=FALSE}
library(tidyverse)
library(openintro)
```


### Pre-exercise

Adding the GGally package.

```{r}
# install.packages("GGally")
library(GGally)
```

Looking at the data from the course package.

```{r}
data(evals)
glimpse(evals)
```

There are 463 rows and 23 columns. 

Evaluating the type of study. 

```{r}
?evals
```




### Exercise 1

Is this an observational study or an experiment? The original research question posed in the paper is whether beauty leads directly to the differences in course evaluations. Given the study design, is it possible to answer this question as it is phrased? If not, rephrase the question.


This is an observational study because the data gathered is not directly interfering with how the data arises and they are merely observing a natural phenomena. 

With an observational study, we cannot make causal claims. We can only provide evidence that an association exists between two variables. 

The question could be rephrased to ask, "is there an association between a professor's attractiveness and their course evaluations?" which could be further broken down into a hypothesis test where: 

$H_o$ There is no relationship between a professor's attractiveness and their course evaluations. 

$H_a$ There is a relationship between a professor's attractiveness and their course evaluations.

It could then be tested to determine the validity of the association between the variables. 

### Exercise 2

Describe the distribution of score. Is the distribution skewed? What does that tell you about how students rate courses? Is this what you expected to see? Why, or why not?


```{r}
hist(evals$score, xlab = "Score", ylab = "Counts", main = "Histogram of Scores")
```

The expectation was a normal distribution (unimodal, bell-shaped around the middle median/mean of 2.5) but this did not occur. The distribution is left skewed which means students rated more positively. This means that either the professors were all great and the random sample resulted in higher numbers of positive ratings by chance, or something else is going on here.  

### Exercise 3

Excluding score, select two other variables and describe their relationship with each other using an appropriate visualization.

Two other variables are the average beauty score and ethnicity. Their distributions are displayed in the boxplots below with the language of the school where the professor received education as a third variable for reference. 

```{r}
ggplot(evals, aes(bty_avg, ethnicity)) + geom_boxplot() + coord_flip() + theme_classic() + labs(x = "Avg Score", y = "Ethnicity", title = "Average Beauty Scores by Ethnicity and Language")
```

We can see the relationship between ethnicity and the average beauty score is about the same between minorities and non-minorities. Their interquartile range is also comparable but minority ethnicities are slightly tighter in distribution than non-minorities. The outer range of nonminorities is wider than minorities. I wonder if this would change if we included new boxplots for professors who went to english and non-english schools.  

```{r}
ggplot(evals, aes(bty_avg, ethnicity, fill = language)) + geom_boxplot() + coord_flip() + theme_classic() + labs(x = "Avg Score", y = "Ethnicity", title = "Average Beauty Scores by Ethnicity and Language")
```

There is certainly a difference in distribution between the ethnicities who's education was received in english and those who received an education that was not in english. For example, minorities that had a non-english education had higher beauty scores on average than those whose education was in english. However, the results are different for non-minorities. This group had an average very close to the average score for all non-minorities that received an education in english. 


### Exercise 4

Replot the scatterplot, but this time use geom_jitter as your layer. What was misleading about the initial scatterplot?

The origional scatterplot is shown below. 

```{r}
ggplot(data = evals, aes(x = bty_avg, y = score)) +
  geom_point()
```

This is the recreation of the scatterplot using geom_jitter.

```{r}
ggplot(data = evals, aes(x = bty_avg, y = score)) +
  geom_jitter()
```
 
```{r eval=FALSE}
?geom_point
?geom_jitter
```
 
 
It appears as though the initial scatterplot was missing some points. Looking into the geom_point and geom_jitter functions, we can see why. The geom_jitter function adds a small amount of random variation to the location of each point making it useful for displaying overlapping points without a discrete overlap as shown in the geom_point scatterplot. 


### Exercise 5

Let’s see if the apparent trend in the plot is something more than natural variation. Fit a linear model called m_bty to predict average professor score by average beauty rating. Write out the equation for the linear model and interpret the slope. Is average beauty score a statistically significant predictor? Does it appear to be a practically significant predictor?


```{r}
ggplot(data = evals, aes(x = bty_avg, y = score)) +
  geom_jitter() +
  geom_smooth(method = "lm", se = FALSE)
```

```{r}
m_bty <- lm(evals$score ~ evals$bty_avg)
summary(m_bty)
```

Let’s see if the apparent trend in the plot is something more than natural variation. Fit a linear model called m_bty to predict average professor score by average beauty rating. Write out the equation for the linear model and interpret the slope. Is average beauty score a statistically significant predictor? Does it appear to be a practically significant predictor?

The equation for the linear model is: 

$$y = 3.88034 + 0.06664$$

Where y is the predicted score of the professor based on x, the average beauty of the professor. 

In this case, the slope is slightly positive at 0.067 and it indicates that as beauty increases the average score of professors on evaluations also increases marginally. 

Based on the low r-squared values (both adjusted and predicted) and the low p-value, it appears as though average beauty is not a practically significant predictor of evaluation score. The low r-squared values indicate that another model may be suited for prediction. Though the low p-value indicates that the relationship between these variables is statistically significant at a level of 0.001. 

### Exercise 6 

Use residual plots to evaluate whether the conditions of least squares regression are reasonable. Provide plots and comments for each one (see the Simple Regression Lab for a reminder of how to make these).


```{r}
hist(m_bty$residuals)
```

A histogram shows the distribution of the residuals to be left skewed and not normally distributed. 

```{r}
plot(m_bty$residuals ~ evals$bty_avg, ylab="Residuals", xlab="Average Beauty", 
main="Rating and Beauty") 
abline(h = 0, lty = 3)  # adds a horizontal dashed line at y = 0
```

This scatterplot fo the residuals also shows the distribution to be left skewed and not normally distributed. The residuals are not constantly variable around the zero line. They are concentrated to the right and might not follow a linear trend. 


### Exercise 7

P-values and parameter estimates should only be trusted if the conditions for the regression are reasonable. Verify that the conditions for this model are reasonable using diagnostic plots.

```{r}
ggplot(data = evals, aes(x = bty_f1lower, y = bty_avg)) +
  geom_point()
evals %>% 
  summarise(cor(bty_avg, bty_f1lower))
evals %>%
  select(contains("bty")) %>%
  ggpairs()
```


```{r}
m_bty_gen <- lm(score ~ bty_avg + gender, data = evals)
summary(m_bty_gen)
```


The conditions for least squares regression are: 

    * Linearity
    * ~ Normal Residuals
    * Constant Variability
    * Independent Observations

These describe that the data should display linear trends, have nearly normal residuals with constant variability, and each observation should be independent of the other observations. 

```{r}
qqnorm(m_bty_gen$residuals)
qqline(m_bty_gen$residuals)
```


We can be confident that the observations are independent of one another. However, given that the residuals are not normal all of these conditions are not met. It may also be the case that the data do not fit a linear model. 


### Exercise 8 

Is bty_avg still a significant predictor of score? Has the addition of gender to the model changed the parameter estimate for gender?

Yes, there was a marginal increase in the intercept and slope but not enough to have a significant change on bty_avg as a significant predictor of score. 

### Exercise 9 

What is the equation of the line corresponding to males? (Hint: For males, the parameter estimate is multiplied by 1.) For two professors who received the same beauty rating, which gender tends to have the higher course evaluation score?


The equation of the line corresponding to males is 

$$score = 3.74734 + 0.07416(b) + 0.17239(m) $$
where b is the b is the average beauty rating and m is the classifier "1" given to males. For females we would use the variable f which is classified as a 0. 

Males tend to have a higher course evaluation score. 
 
### Exercise 10 

Create a new model called m_bty_rank with gender removed and rank added in. How does R appear to handle categorical variables that have more than two levels? Note that the rank variable has three levels: teaching, tenure track, tenured.

```{r}
m_bty_rank <- lm(score ~ bty_avg + rank, data = evals)
summary(m_bty_rank)
```

Since only two categorical variables appear in the results of the linear regression, I would assume that R is purposefully excluding one of the three categories. In this case, the missing category is teaching. 

### Exercise 11 

Which variable would you expect to have the highest p-value in this model? Why? Hint: Think about which variable would you expect to not have any association with the professor score.

The variable that should be the least likely to have any association with the professor score is the cls_level. In this case, the level of the class as upper or lower should not have any association with the professor score. Additionally, there is often overlap in upper and lower level courses for all students.  

### Exercise 12

Check your suspicions from the previous exercise. Include the model output in your response.

```{r}
m_full <- lm(score ~ rank + gender + ethnicity + language + age + cls_perc_eval 
             + cls_students + cls_level + cls_profs + cls_credits + bty_avg 
             + pic_outfit + pic_color, data = evals)
summary(m_full)
```

My assumption was incorrect. Cls_profs had the highest p-value and thus the least likely association with score. This categorical variable observes whether the course had multiple professors teaching sections in the course or if there was only a single professor teaching all sections for the course. It turns out that this has little to no influence on the professor score. 


### Exercise 13 

Interpret the coefficient associated with the ethnicity variable.

Holding all other variables constant, we would expect that a non-minority professor would rate about 0.1235 points higher on their evaluations. However, given the large p-value the results would not be significant as there is a poor association between the professor's ethnicity and the professor's score. 

### Exercise 14 

Drop the variable with the highest p-value and re-fit the model. Did the coefficients and significance of the other explanatory variables change? (One of the things that makes multiple regression interesting is that coefficient estimates depend on the other variables that are included in the model.) If not, what does this say about whether or not the dropped variable was collinear with the other explanatory variables?

```{r}
m_full_drop1 <- lm(score ~ rank + gender + ethnicity + language + age + cls_perc_eval 
             + cls_students + cls_level + cls_credits + bty_avg 
             + pic_outfit + pic_color, data = evals)
summary(m_full_drop1)
```

Yes, there were slight changes in the the coefficients and significance of the other explanatory variables. 

### Exercise 15 

Using backward-selection and p-value as the selection criterion, determine the best model. You do not need to show all steps in your answer, just the output for the final model. Also, write out the linear model for predicting score based on the final model you settle on.

```{r}
m_full_best <- lm(score ~ gender 
                  + ethnicity 
                  + age 
                  + language
                  + cls_perc_eval 
              + cls_credits 
              + bty_avg 
             + pic_color, data = evals)
summary(m_full_best)
```

The best linear model for predicting score based on backward-selection and p-value significance is: 

$$score = 3.771922+0.207112(gender)+0.167872(ethnicity)-0.006046(age) - 0.206178(language) +0.004656(cls_perc_eval)+ 0.505306(cls_credits)+ 0.051069(bty_avg)-0.190579 (pic_color)$$

### Exercise 16 

Verify that the conditions for this model are reasonable using diagnostic plots.

The conditions are:

    * Linearity
    * ~ Normal Residuals
    * Constant Variability
    * Independent Observations

We can evaluate using the diagnostic plots of the final regression. 

```{r}
m_bty_final <- lm(score ~ ethnicity 
                  + gender 
                  + language 
                  + age 
                  + cls_perc_eval 
              + cls_credits 
              + bty_avg 
              + pic_color, data = evals)
qqnorm(m_bty_final$residuals)
qqline(m_bty_final$residuals)
```

There is a slight deviation from normal towards both ends of this plot, but it is still nearly normal. It is also reasonbly linear. 


```{r}
plot(m_bty_final$residuals ~ evals$bty_avg, xlab = "Average Beauty Score", ylab = "Final Residuals", main = "Residuals of Final Average Beauty Scores")
abline(h = 0, lty = 3)
```

The residuals appear relativly constant in their variability. They are also nearly normal and reasonbly spread around the zero line. 


```{r}
hist(m_bty_final$residuals) 
```
In this histogram we can see a slight left skew, however, it might not be enough to worry about. Assuming that each observation is independent since each student made their own decisions about a single professor just once and separately without influence of another's evaluation, then these conditions have been met. 


### Exercise 17 

The original paper describes how these data were gathered by taking a sample of professors from the University of Texas at Austin and including all courses that they have taught. Considering that each row represents a course, could this new information have an impact on any of the conditions of linear regression?

Yes, this new information would violate the condition independent observations. 



### Exercise 18 

Based on your final model, describe the characteristics of a professor and course at University of Texas at Austin that would be associated with a high evaluation score.

Based on the final model, the characteristics of a professor and course at the University of Texas at Austin that would be associated with a high evaluation score are: 

    * Male
    * Relativly young
    * Non-minorities
    * Have english education
    * Teach one-credit courses
    * Provided color photos
    * High average beauty score
    * High % of students complete evaluations


### Exercise 19 

Would you be comfortable generalizing your conclusions to apply to professors generally (at any university)? Why or why not?


No, I would not be comfortable generalizing these conclusions to apply to professors generally at any university. To start, although this model was selected for statistical significance, it was entirely based on the population of students at one unversity in one place for a specific set of time. This will change over-time. Our ability to predict the values is also very low, as evident in the low R-squared and Adjusted R-sqaured values. There are also many cultural norms and perceived benefits of professors in Austin that will not be seen as a benefit to students in another country. 

For example, if this were applied at the University of Osmani in Hyderabad, India, the model would likely render itself ineffective even with huge samples of the population. This is a minority dominated, non-english based, male-dominated school. Professors are often much older and the cultural norms of what makes a person beautiful at this University take a very different perspective than what students at the University of Austin would say. The variables in the model would no longer hold normal and conditions for regression would no longer be met. 




...

