---
title: "HW4"
author: "Zachary Palmore"
date: "4/17/2021"
output: pdf_document
subtitle: "Business Analytics and Data Mining"
header-includes: 
- \newcommand{\bcenter}{\begin{center}}
- \newcommand{\ecenter}{\end{center}}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\bcenter

# Assignment 4

\ecenter

## Purpose 

In this homework assignment, we will explore, analyze and model a data set containing approximately 8000 
records representing a customer at an auto insurance company. Each record has two response variables. The 
first response variable, TARGET_FLAG, is a 1 or a 0. A “1” means that the person was in a car crash. A zero 
means that the person was not in a car crash. The second response variable is TARGET_AMT. This value is zero 
if the person did not crash their car. But if they did crash their car, this number will be a value greater than zero.

Our objective is to build multiple linear regression and binary logistic regression models on the training data 
to predict the probability that a person will crash their car and also the amount of money it will cost if the person 
does crash their car. We can only use the variables given (or variables derived from the variables 
provided). Below is a short description of the variables of interest in the data set:



```{r}
# short descriptions of variables as table from matrix
```



## Introduction

```{r}
tdata <- read.csv("https://raw.githubusercontent.com/palmorezm/msds/main/621/HW4/insurance_training_data.csv")
edata <- read.csv("https://raw.githubusercontent.com/palmorezm/msds/main/621/HW4/insurance-evaluation-data.csv")
head(tdata)
```

## Data Exploration

Describe the size and the variables in the insurance training data set. Consider that too much detail will cause a 
manager to lose interest while too little detail will make the manager consider that you aren’t doing your job. Some 
suggestions are given below. Please do NOT treat this as a check list of things to do to complete the assignment. 
You should have your own thoughts on what to tell the boss. These are just ideas.
a. Mean / Standard Deviation / Median
b. Bar Chart or Box Plot of the data
c. Is the data correlated to the target variable (or to other variables?)
d. Are any of the variables missing and need to be imputed “fixed”?


## Data Preparation

Describe how you have transformed the data by changing the original variables or creating new variables. If you
did transform the data or create new variables, discuss why you did this. Here are some possible transformations.
a. Fix missing values (maybe with a Mean or Median value)
b. Create flags to suggest if a variable was missing
c. Transform data by putting it into buckets
d. Mathematical transforms such as log or square root (or use Box-Cox)
e. Combine variables (such as ratios or adding or multiplying) to create new variables

## Model Building

Using the training data set, build at least two different multiple linear regression models and three different binary 
logistic regression models, using different variables (or the same variables with different transformations). You 
may select the variables manually, use an approach such as Forward or Stepwise, use a different approach such 
as trees, or use a combination of techniques. Describe the techniques you used. If you manually selected a 
variable for inclusion into the model or exclusion into the model, indicate why this was done.

Discuss the coefficients in the models, do they make sense? For example, if a person has a lot of traffic tickets, 
you would reasonably expect that person to have more car crashes. If the coefficient is negative (suggesting that 
the person is a safer driver), then that needs to be discussed. Are you keeping the model even though it is counter 
intuitive? Why? The boss needs to know.

## Model Selection

Decide on the criteria for selecting the best multiple linear regression model and the best binary logistic regression 
model. Will you select models with slightly worse performance if it makes more sense or is more parsimonious? 
Discuss why you selected your models. 
For the multiple linear regression model, will you use a metric such as Adjusted R2
, RMSE, etc.? Be sure to 
explain how you can make inferences from the model, discuss multi-collinearity issues (if any), and discuss other 
relevant model output. Using the training data set, evaluate the multiple linear regression model based on (a) 
mean squared error, (b) R2
, (c) F-statistic, and (d) residual plots. For the binary logistic regression model, will you 
use a metric such as log likelihood, AIC, ROC curve, etc.? Using the training data set, evaluate the binary logistic 
regression model based on (a) accuracy, (b) classification error rate, (c) precision, (d) sensitivity, (e) specificity, (f) 
F1 score, (g) AUC, and (h) confusion matrix. Make predictions using the evaluation data set.















