---
title: "Project 1"
author: "Group 3"
date: "6/19/2021"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Given an unknown data source with several periods, we attempt to predict the next 140 based on the total 1622 periods in this data set. Our predictions will be fine-tuned to reduce the mean absolute percentage error (MAPE) as much as possible. The packages we will be using and the data, with its first five rows, is shown below. 

```{r}
# Packages
library(tidyverse)
library(forecast)
library(MLmetrics)
library(openxlsx)

# Data source
data <- read.csv("https://raw.githubusercontent.com/palmorezm/msds/main/624/Projects/project1data.csv")
data <- data %>% 
  rename(SeriesIND = ï..SeriesInd) 
head(data, 5)
```

Our results will be in an excel spreadsheet containing one sheet (or tab) for each group. Predictions will be made for each of these groups based on the most appropriate model developed in this analysis. Each sheet will also have the variable *SeriesIND* listed ascending order for review. 

```{r}
# Grouping
S01 <- data %>% 
  filter(group == "S01")
S02 <- data %>% 
  filter(group == "S02")
S03 <- data %>%
  filter(group == "S03")
S04 <- data %>% 
  filter(group == "S04")
S05 <- data %>% 
  filter(group == "S05")
S06 <- data %>% 
  filter(group == "S06")

# Imputation by function - missing something? lapply/sapply may work 
soximp <- function(df){
  for (i in colnames(df)){
    if (sum(is.na(df[[i]])) !=0){
      df[[i]][is.na(df[[i]])] <- median(df[[i]], na.rm=TRUE)
    }
  }
}

# Imputation loops for each group by median 
for (i in colnames(S01)){
  if (sum(is.na(S01[[i]])) != 0){
    S01[[i]][is.na(S01[[i]])] <- median(S01[[i]], na.rm = TRUE)
  } 
}
for (i in colnames(S02)){
  if (sum(is.na(S02[[i]])) != 0){
    S02[[i]][is.na(S02[[i]])] <- median(S02[[i]], na.rm = TRUE)
  } 
}
for (i in colnames(S03)){
  if (sum(is.na(S03[[i]])) != 0){
    S03[[i]][is.na(S03[[i]])] <- median(S03[[i]], na.rm = TRUE)
  } 
}
for (i in colnames(S04)){
  if (sum(is.na(S04[[i]])) != 0){
    S04[[i]][is.na(S04[[i]])] <- median(S04[[i]], na.rm = TRUE)
  } 
}
for (i in colnames(S05)){
  if (sum(is.na(S05[[i]])) != 0){
    S05[[i]][is.na(S05[[i]])] <- median(S05[[i]], na.rm = TRUE)
  } 
}
for (i in colnames(S06)){
  if (sum(is.na(S06[[i]])) != 0){
    S06[[i]][is.na(S06[[i]])] <- median(S06[[i]], na.rm = TRUE)
  } 
}
```

#Create fit data
```{r}
S01_fitVar01_data = S01[1:1622,c('SeriesIND','Var01')]
S01_fitVar02_data = S01[1:1622,c('SeriesIND','Var02')]
S01_fitVar05_data = S01[1:1622,c('SeriesIND','Var05')]

S02_fitVar01_data = S02[1:1622,c('SeriesIND','Var01')]
S02_fitVar02_data = S02[1:1622,c('SeriesIND','Var02')]
S02_fitVar05_data = S02[1:1622,c('SeriesIND','Var05')]

S03_fitVar01_data = S03[1:1622,c('SeriesIND','Var01')]
S03_fitVar02_data = S03[1:1622,c('SeriesIND','Var02')]
S03_fitVar05_data = S03[1:1622,c('SeriesIND','Var05')]

```

## Analysis
```{r}
fit_S01_Var01 = auto.arima(S01_fitVar01_data[,'Var01'])
fit_S01_Var02 = auto.arima(S01_fitVar02_data[,'Var02'])
fit_S01_Var05 = auto.arima(S01_fitVar05_data[,'Var05'])

fit_S02_Var01 = auto.arima(S02_fitVar01_data[,'Var01'])
fit_S02_Var02 = auto.arima(S02_fitVar02_data[,'Var02'])
fit_S02_Var05 = auto.arima(S02_fitVar05_data[,'Var05'])

fit_S03_Var01 = auto.arima(S03_fitVar01_data[,'Var01'])
fit_S03_Var02 = auto.arima(S03_fitVar02_data[,'Var02'])
fit_S03_Var05 = auto.arima(S03_fitVar05_data[,'Var05'])

```

## Prediction
```{r}
predict_S01_Var01 = fit_S01_Var01 %>% forecast(h=140) %>% data.frame
predict_S01_Var02 = fit_S01_Var02 %>% forecast(h=140) %>% data.frame
predict_S01_Var05 = fit_S01_Var05 %>% forecast(h=140) %>% data.frame

predict_S02_Var01 = fit_S02_Var01 %>% forecast(h=140) %>% data.frame
predict_S02_Var02 = fit_S02_Var02 %>% forecast(h=140) %>% data.frame
predict_S02_Var05 = fit_S02_Var05 %>% forecast(h=140) %>% data.frame

predict_S03_Var01 = fit_S03_Var01 %>% forecast(h=140) %>% data.frame
predict_S03_Var02 = fit_S03_Var02 %>% forecast(h=140) %>% data.frame
predict_S03_Var05 = fit_S03_Var05 %>% forecast(h=140) %>% data.frame

```

#Write predictions to csv and Excel
```{r}
predictions_S123_v125_df = data.frame(predict_S01_Var01[1],predict_S01_Var02[1], predict_S01_Var05[1],
predict_S02_Var01[1],predict_S02_Var02[1],predict_S02_Var05[1],predict_S03_Var01[1],predict_S03_Var02[1],
predict_S03_Var05[1])

colnames(predictions_S123_v125_df) <- c('S01_Var01', 'S01_Var02', 'S01_Var05', 'S02_Var01', 'S02_Var02', 'S02_Var05', 'S03_Var01', 'S03_Var02', 'S03_Var05')

write.csv(x = predictions_S123_v125_df, file='predictions_S123_V125_Ken.csv')
write.xlsx(predictions_S123_v125_df, 'predictions_S123_V125_Ken.xlsx')
```

