---
title: "Project 2 - Brazilian Climate Data"
author: "Zachary Palmore"
date: "9/21/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


### Objective

In this part of project 2, I will be tidying and briefly analyzing climate data from the National Meteorological Institute of Brazil (IMET). It contains hourly surface weather conditions for the Southeastern states of Rio de Janeiro, São Paulo, and Minas Gerais e Espirito Santo. What we would like to know is;

  * Max, min, and mean of temperatures for each station     
  * Variance in temperature over the study area (SE Brazil) 
  * Relationship between temperature and humidity  

To get those answers, the data needs to be cleaned and tidyed before we can make use of it. 

```{r message=FALSE}
library(readr)
library(tidyverse)
```

### Importing Data

The link provided was the most streamlined way to get the data without needing permission. It appeared to be 

```{r}
surfaceweather <- read.delim("C:/Users/Owner/Downloads/surfaceweather.txt", header = TRUE, quote="", sep = "\t", dec = ".", fill = TRUE, comment.char = "")
# surfaceweather <- read_delim("C:/bigdata/surfaceweather.txt", 
#    "\t", escape_double = FALSE, trim_ws = TRUE)
# surfaceweather <- read_delim("C:/bigdata/surfaceweather.txt", 
#    "\t", escape_double = FALSE, trim_ws = TRUE)
head(surfaceweather)
```



The shear amount of data makes it difficult to find what we need. To solve this, I will select out which columns are needed. 



```{r}
climate <- subset(surfaceweather, select = c("wsid", 
                                             "mdct",
                                             "temp",
                                             "hmdy"), na.rm = TRUE)
head(climate)
```



```{r}
summary(climate)
```



```{r}
var(climate$hmdy)
```

```{r}
wsid178 <- climate %>%
  filter(wsid == 178)
unique(climate$wsid)
  
```

```{r}
sel_climate <- climate[1:800,1:4]
sel_climate <- sel_climate %>%
  filter(temp > 0)
```




```{r}
climate <- climate %>%
  filter(temp > 0) %>%
  filter(hmdy > 0)
```



```{r}
ggplot(data = climate, aes(x = temp, y = hmdy)) + 
  geom_point(colour = "black", 
             size = 1.5, 
             shape = 46, 
             alpha = 1/100,
             na.rm = TRUE) +
  xlab("Temperature") +
  ylab("Humidity") +
  labs( title = "Title", subtitle = "Subtitle",
        caption = "caption") +
  geom_smooth(method = lm, formula = y ~ x) +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5), plot.caption = element_text(hjust = 0.5))
```



