---
title: "Project 2 - Uber "
author: "Zachary Palmore"
date: "9/21/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


### Objective

In this part of project 2, I will be tidying and briefly analyzing data from Uber. It contains the quarterly earnings before interest, taxes, depreciation, and amortization (EBITDA). What we would like to know is;

  * Max, min, and mean of temperatures for each station     
  * Variance in temperature over the study area (SE Brazil) 
  * Relationship between temperature and humidity  

To get those answers, the data needs to be cleaned and tidied before we can make use of it. 

```{r message=FALSE}
require(tidyverse)
require(magick)
require(tesseract)
require(reshape2)
```



### Importing Data

The image provided contains all the information needed.  It appeared to be 


```{r}
# preprocessing
uber <- image_read("uber.png") %>% 
  image_negate() %>% 
  image_transparent("transparent", fuzz=10) %>% 
  image_background("white") %>%
  image_morphology(method = "Thinning", kernel = "Rectangle") %>%
  image_crop(geometry_area(-10, -40, -100, -20)) %>%
  image_chop(geometry_area(
    width = 14, 
    height = 125, 
    x_off = 1, 
    y_off = 1.5))

uber
```

```{r}
uber <- uber %>% 
  image_ocr()
uber
uber <- read_delim(uber, col_names = FALSE, " ", comment = "", escape_double = TRUE)
```

```{r}

```


```{r}

```



```{r}
uber <- readPNG("uber.png")
# f <- system.file("uber.png", package="raster")
ub <- system.file()
rasterToPoints(uber)
```


```{r}
r <- raster("/GitHub/msdsdata607/Projects/Project 2/uber.png")
x <- rasterToPoints(r)
extract(x,2)
```



```{r}
f <- system.file("/GitHub/msdsdata607/Projects/Project 2/uber.png", package="raster")
r <- raster(f)
?raster
x <- as.data.frame(r)


f <- system.file("uber.png", package = "raster")
raster(f)


head(x)
```




```{python}
print("hello")
```




