---
title: "Project 1"
author: "Zachary Palmore"
date: "9/17/2020"
output: 
  pdf_document:
        extra_dependencies: ["geometry", "multicol", "multirow"]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Chess Tournament Text Conversion

```{r}
library(tidyverse)
library(DataCombine)
```



### Directions

In this project, you’re given a text file with chess tournament results where the information has some structure. Your job is to create an R Markdown file that generates a .CSV file (that could for example be imported into a SQL database) with the following information for all of the players:

Player’s Name, Player’s State, Total Number of Points, Player’s Pre-Rating, and Average Pre Chess Rating of Opponents

For the first player, the information would be:
Gary Hua, ON, 6.0, 1794, 1605

1605 was calculated by using the pre-tournament opponents’ ratings of 1436, 1563, 1600, 1610, 1649, 1663, 1716, and dividing by the total number of games played.

___

\clearpage


### Step 1


```{r include = FALSE}
#checking to ensure Gary Hua's rating is a mean with the directions
avg_rating <- c(1436, 1563, 1600, 1610, 1649, 1663, 1716)
mean(avg_rating)
```


Importing the text file 

```{r}
library(readr)
tournamentinfo <- read_delim("https://raw.githubusercontent.com/palmorezm/msdsdata607/master/tournamentinfo.txt", 
    "|", escape_double = FALSE, comment = "--", 
    trim_ws = TRUE)
head(tournamentinfo)
```


### Solution


```{r}
tournamentinfo <- read_lines("https://raw.githubusercontent.com/palmorezm/msdsdata607/master/tournamentinfo.txt")
hyphens <- str_detect(tournamentinfo, '^[-]{2,}$') 
tournamentinfo <- tournamentinfo[!hyphens == "TRUE"]
tournamentinfo <- str_remove_all(tournamentinfo, "[WDL]") 
tournamentinfo <- str_replace_all(tournamentinfo, "[|/]",",") 
empty <- c("") 
for (i in seq(1, length(tournamentinfo)-1, by = 2)){
   empty <- c(empty, paste(tournamentinfo[i], tournamentinfo[i+1], sep = "", collapse = NULL))
}
TournamentResults <- as.data.frame(do.call(rbind, strsplit(empty, ",")), stringsAsFactors = FALSE)
TournamentResults
```

```{r}
# str_match_all(tournamentinfo, "R:\\s*(.*?)\\s*->")
Preratings <- str_extract_all(tournamentinfo, "R:\\s*(.*?)\\s*->")
Preratings <- str_remove_all(Preratings, "R:")
Preratings <- str_remove_all(Preratings, "->")
Preratings <- str_remove_all(Preratings, "P\\d")
Preratings <- str_match_all(Preratings, "\\d+")
Preratings <- str_remove_all(Preratings, "0")
Preratings <- as.data.frame(Preratings)
Preratings <- as.numeric(as.character(unlist(Preratings[[1]])))
Preratings <- as.data.frame(Preratings)
Preratings <- na.omit(Preratings)
```


```{r}

view(Preratings)
rbind.data.frame(TournamentResults, Preratings)
```




```{r}
view(tournamentinfo)
view(TournamentResults)
tournamentinfo
str_subset(tournamentinfo, "\\d{4} \\d{4}")
# Character 0
str_subset(tournamentinfo, "R:\\d{4}")
# Character 0
str_subset(tournamentinfo, "R: \\d{4} \\d{4}")
# Character 0 
str_subset(tournamentinfo, "\\d{4}")
# character 0
str_detect(tournamentinfo, "R: \\d{4} \\d{4}")
# All False
str_detect(tournamentinfo, "\\d{4} \\d{4}")
# All False
str_detect(tournamentinfo, "\\d{4}")
# Many True (too many)
str_subset(TournamentResults, "R: \\d{4}")
# Pull everything to the right of R: including R:
str_remove_all(tournamentinfo, "R:")
# Removed  "R:" and left everything before and after it
str_subset(tournamentinfo, ())
```





```{r}
snippet = rep(0, length(tournamentinfo))
 for (i in 1:length(tournamentinfo))
     {
          # extract the initial position
          initial.position = gregexpr("R:", tournamentinfo[i])[[1]][1] + 1
  
          # extract the final position
          final.position = gregexpr("->", tournamentinfo[i])[[1]][1] - 1
 
          # extract the substring between the initial and final positions, inclusively
          snippet[i] = substr(tournamentinfo[i], initial.position, final.position)
     }
 view(snippet)
}
```

```{r}
str_remove_all(snippet, "P\\d{2}")
view(snippet)

# Results <- cbind(TournamentResults, snippet) 
# view(Results)
```





___
\clearpage

### Conclusion

After performing the conversion we were able to create a spreadsheet with the following variables for each player:

Name     | State | Total | Pre-rating | Avg-Rate
---------|-------|-------|------------|----------
Gary Hua | ON    | 6.0   | 1794       | 1605
---------|-------|-------|------------|----------

This can be shown in the following data frame with the first five players selected alongside their statistics. 

```{r}
# data frame with first 5 players stats selected
```


The data frame can be converted to a spreadsheet (specifically a ".csv") with the following chunk. 

```{r}
# Converting to a .csv from data frame
# write.csv(Your DataFrame,"Path where you'd like to export the DataFrame\\File Name.csv", row.names = FALSE)
# Source https://datatofish.com/export-dataframe-to-csv-in-r/
write.csv(TournamentResults, "C:\\Users\\Zachary Palmore\\Desktop\\TournamentResults.csv", row.names = FALSE) 
```

Data is exported to my desktop but could be pulled into anyone's desktop or an SQL database. 


